---
title: "Sediment toxicity analysis"
author: ""
output: 
  html_document:
    code_folding: hide
self_contained: yes
---

```{r, echo = F}
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, fig.path = 'figs/', dev.args = list(bg = 'transparent'), eval = F)
```

```{r setup, eval = T}
library(tidyverse)
library(GGally)
library(vegan)
library(ggdendro)
library(dendextend)
library(FactoMineR)
library(ggord)
library(mapview)
library(sf)

data(dat)

locs <- dat %>% 
  select(station, lon, lat)
evvr <- dat %>% 
  select(-toc, -fine, -tn, -depth, -lat, -lon)
```

The sediment toxicity dataset includes measurements at 155 coastal sites as part of the Bight 2008 sampling.  Twenty-eight measurements are available for each site: total organic carbon (`toc`), total nitrogen (`tn`), percent fine sediment (`fine`), amphipod survival (`amph_surv`), embryonic survival (`embr_surv`), and twenty-three metals.  Earlier analyses indicated that `tn`, `fine`, and `toc` were not related to survivability and these measurements were not evaluated further.  This exploratory analysis was conducted to 1) evaluate correlations among measurements across the sites and 2) identify if sites could be characterized by similarities in characteristics.  The objective was to identify patterns among sites that may explain differences in survival rates or by location.  

This plot shows relationships among raw variables - outliers are present and variables occur at different scales.
```{r pairplot, fig.height = 9, fig.width = 9, eval = T}
pairs(evvr[, -1], gap = 0, cex = 0.5, col = 'grey')
```

This plot is identical to the above but shows the variables after removing outliers and standardizing the scales (substract the mean and divide by variance).  This 'clean' dataset was used for all analyses.
```{r pairplot_cln, fig.height = 9, fig.width = 9, eval = T}
dat_frm <- evvr %>%
  data.frame %>%
  filter(amph_surv > 30) %>%
  filter(Hg < 5) %>% 
  filter(Cr < 100) %>% 
  filter(Ba < 600) %>% 
  filter(Sb < 1.3) %>% 
  filter(Ag < 2) %>% 
  mutate(
    embr_surv = ifelse(is.na(embr_surv), median(embr_surv, na.rm = TRUE), embr_surv), 
    Al = ifelse(is.na(Al), median(Al, na.rm = TRUE), Al),
    Sb = ifelse(is.na(Sb), median(Sb, na.rm = TRUE), Sb),
    As = ifelse(is.na(As), median(As, na.rm = TRUE), As)
  ) %>%
  column_to_rownames('station') %>%
  decostand(method = 'standardize')

pairs(dat_frm, gap = 0, cex = 0.5, col = 'grey')
```

The `kmeans` algorithm was used to assign sites to similar groups.  Group assignment using this algorithm is based on an iterative process where sites are assigned to minimize the within-group averages of the environmental variables.  The number of groups was chosen to maximize the separation in multivariate space (see the PCA biplot below) and based on spatial groupings of the data (see the map below).  The pairs plot below colors the sites by these groupings. 
```{r dendro, fig.width = 9, fig.height = 9, eval = T}
set.seed(1234)
ngrps <- 4
grps <- kmeans(dat_frm, centers = ngrps, nstart = 10, iter.max = 30) %>% 
  .$cluster
cols <- mapviewGetOption("vector.palette")(ngrps)

colgrp <- data.frame(
  grps = c(1:ngrps), 
  cols = cols, 
  stringsAsFactors = F
)

colprs <- left_join(
    data.frame(grps = grps), colgrp, by = 'grps'
  )

pairs(dat_frm, gap = 0, cex = 0.5, col = as.character(colprs$cols))
```

The boxplots show the individual distributions of the environmental variables by the groups.
```{r, fig.height = 9, fig.width = 9, eval = T}
toplo <- dat_frm %>% 
  mutate(grps = grps) %>% 
  left_join(colgrp, by = 'grps') %>% 
  mutate(grps = factor(grps)) %>% 
  gather('var', 'val', -grps, -cols)
mythm <- theme_bw(base_family = 'serif') +
  theme(
  legend.position = 'none', 
  strip.background = element_blank(), 
  plot.margin = margin(0, 2, 0, 2, "pt")
) 

colord <- unique(toplo[, c('grps', 'cols')])

ggplot(toplo, aes(x = grps, y = val)) + 
  geom_boxplot(outlier.size = -1) + 
  geom_jitter(colour = toplo$cols, alpha = 0.3, width = 0.15) + 
  facet_wrap(~var, ncol = 5) + 
  mythm + 
  scale_y_continuous('Standardized measurements')
  
```

The site groupings from the kmeans clusters and variance among the measured variables were also evaluated using principal components analysis.  The ordination plot shows sites with similar characteristics closer to each other in multivariate space.  In addition, the measured variables that are related to site groupings are also plotted as two-dimensional vectors.  The point cloud groupings in the plot were identified from the kmeans clustering. Axis one explains a gradient in toxicity from low to high, axis two shows which groups of toxicity variables are dominant when measurements are high (separating the 'high' groups 1 and 4), and axis 3 somewhat shows which variables were not important for distinguishing between groups 1 and 4. 
```{r ordiplot, fig.height = 7, fig.width = 7, eval = T}

ppp <- PCA(dat_frm, scale.unit = F, graph = F)

mythm <- theme_bw(base_family = 'serif') +
  theme(
  legend.position = 'none', 
  strip.background = element_blank(), 
  plot.margin = margin(0, 2, 0, 2, "pt")
) 

p1 <- ggord(ppp, grp_in = as.character(grps), vec_ext = 9, alpha = 0.8, size = 2, txt = 4, arrow = 0.2, cols = cols) + 
  mythm
p1

p2 <- ggord(ppp, grp_in = as.character(grps), axes = c('2', '3'), vec_ext = 9, alpha = 0.8, size = 2, txt = 4, arrow = 0.2, cols = cols) + 
  mythm
p2

p3 <- ggord(ppp, grp_in = as.character(grps), axes = c('1', '3'), vec_ext = 9, alpha = 0.8, size = 2, txt = 4, arrow = 0.2, cols = cols) + 
  mythm
p3
```

Finally, this map shows the site locations that are color-coded by the clusters.  It can be used to evaluate spatial differences in the groupings.  
```{r, eval = T}
prstr <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0" 

toplo <- dat_frm %>% 
  rownames_to_column('station') %>% 
  mutate(station = as.numeric(station)) %>% 
  left_join(locs, by = 'station') %>% 
  mutate(Groups = grps) %>% 
  filter(!is.na(lon)) %>% 
  st_as_sf(coords = c('lon', 'lat'), crs = prstr)

mapview(toplo, zcol = 'Groups')
```

Conclusions: 

* The toxicity measurements show a clear gradient from high to low across the sites.
* The sites vary in which toxicity measurements are highest.  
* The four groupings identified with kmeans clustering can be described as follows.  The yellow and purple groups are both high in toxicity but differ in which variables are dominant.
     * green (Group 3): low for all toxicity measurements, 
     * blue (Group 2): intermediate or not explained by any of the variables
     * yellow (Group 4): high in toxicity
     * purple (Group 1): high in toxicity
* Amphipod and embryonic survivability were not strongly associated with any particular group, although some associations were observed. In particular:
     * Embryonic survivability was somewhat lower in the yellow group
     * Amphipod survivability was somewhat higher in the green group, somewhat lower in the purple group
* Spatial patterns were clear among the groupings:
     * The yellow group was most common in San Diego Bay
     * The purple group was most common in the port of LA
* To do:
     * Which toxicity variables are driving the group differences
     * Which toxicity variables are most strongly associated with survivorship
